---
editor_options: 
  chunk_output_type: console
---

```{r}
library(tidyverse)
library(patchwork)
```


# data preparation
Due to rounding issues during the sequencing data simulation with ART, we aren't guaranteed an exact total of the targeted 20M reads per sample. The first step is to read in the data summary file to calculate the per-sample total depth.

```{r}
#if (!file.exists("benchmarking/data_summary.tsv")) system("scp lilac:/data/brinkvd/users/watersn/spikequant/benchmarking/data_summary.tsv benchmarking/data_summary.tsv")
#if (!file.exists("benchmarking/results/merged_benchmarking_coverage.tsv")) system("scp lilac:/data/brinkvd/users/watersn/spikequant/benchmarking/results/merged_benchmarking_coverage.tsv benchmarking/results/merged_benchmarking_coverage.tsv"


spikes <- c("Haloarcula_hispanica", "Salinibacter_ruber", "Trichoderma_reesei")
spikes_pri <- paste0(spikes, "_primary") 
spikes_pri_masked <- paste0(spikes, "_primary_masked") 
# read in the data summary to get the actual totals 
data_summary_most <- read.csv("benchmarking/data_summary.tsv", sep="\t") %>%
  #filter(!grepl("spike0.00000", file)) %>% 
  mutate(
    true_count=as.numeric(gsub(",", "", num_seqs)),
    sample=gsub(".*?_(\\d+_depth.*?_spike.*?)_.*", "\\1", file),
    Genome = gsub(".*?_(\\d+_depth.*?_spike.*?)_(.*).fasta.*", "\\2", file),
    is_spike =Genome %in% c(spikes, spikes_pri)
  ) %>% 
  group_by(sample,is_spike, Genome) %>% 
  summarize(true_count = sum(true_count)) %>%
  ungroup() %>%
  group_by(sample) %>% 
  mutate(total = sum(true_count)) %>%
  ungroup() %>% 
  filter( is_spike)%>% 
  select(sample, Genome, true_count, total) %>% 
  distinct()
# parse the ones lacking any spikes so we still have totals
# the left join with the dummy data makes the merging below more straightforward
data_summary_nospike <- read.csv("benchmarking/data_summary.tsv", sep="\t") %>% 
  filter(grepl("spike0.00000", file)) %>% 
  mutate(
    true_count=as.numeric(gsub(",", "", num_seqs)),
    sample=gsub(".*?_(\\d+_depth.*?_spike.*?)_.*", "\\1", file),
  ) %>% 
 group_by(sample) %>% 
  summarize(total = sum(true_count)) %>% ungroup() %>%
  mutate(true_count=0) %>% 
  left_join(data.frame(true_count=rep(0, length(spikes)), Genome=spikes), relationship= "many-to-many")

data_summary <- bind_rows(data_summary_most,data_summary_nospike) %>% 
  rename("genome_base" = "Genome")
```


Next, because we masked some regions of the genome for the rrna masked set, we need to determine the adjusted genome size by tallying how many bases from the overall length we lost in the process
```{r}
# filter only primary contigs, and rename assemblies
rrna_depletion <- read.csv("benchmarking/spikes/rrna_masking.bed", sep="\t", header=FALSE, col.names = c("seq", "source", "feature", "start", "stop", "score", "strand",  "frame", "dets")) %>% 
  mutate(Genome = case_when(
    seq == "NZ_RQWK01000001.1" ~ "Haloarcula_hispanica_primary_masked",
    seq == "CP000159.1" ~ "Salinibacter_ruber_primary_masked",
    TRUE ~ "whocares"
  )) %>% 
  mutate(feature_len = stop-start) %>% 
  filter(Genome %in% spikes_pri_masked) %>% 
  group_by(Genome) %>% 
  summarize(total_rrna_length = sum(feature_len))
```



```{r}
tmp2 <- read.csv("benchmarking/v2/all_coverage.tsv", sep="\t")  %>% 
  rename(rawsample=sample) %>%
  mutate(
    rep=gsub("(.*?)_.*quanttype(.*)$", "\\1", rawsample),
    quantref=gsub(".*quanttype(.*)$", "\\1", rawsample),
    sample=gsub("(.*?)_offtarget.*", "\\1", rawsample),
    dbtype=gsub("(.*?)_offtarget(.*?)_.*", "\\2", rawsample),
    aligner=gsub("(.*?)_mapper(.*?)_.*", "\\2", rawsample),
    target=as.numeric(gsub(".*_spike(.*?)_.*", "\\1", rawsample)),
    is_spike =Genome %in% c(spikes, spikes_pri, spikes_pri_masked)
  ) %>%
  mutate(genome_base = gsub("_primar.*", "", Genome )) %>% 
  left_join(data_summary, by = c("sample", "genome_base")) %>% 
  group_by(genome_base) %>% 
  mutate(full_genome = max(Length)) %>% 
  ungroup() %>% 
  left_join(rrna_depletion) %>% 
  mutate(adjusted_length = ifelse(is.na(total_rrna_length), Length, Length-total_rrna_length)) %>%
  mutate(counts = (Read_Count * (full_genome/100000))/(adjusted_length/100000)) %>% 
  select(rep, sample, dbtype, quantref, aligner, is_spike, target, Genome, Length, genome_base, full_genome, Read_Count, counts, true_count, total,adjusted_length) %>% 
  mutate(count_deviation_percent = (counts-true_count)/(true_count)) %>% 
  mutate(count_percent = (true_count) /counts,
         count_diff = counts - true_count,
         count_percent = ifelse(count_diff == 0, 1, 
                                ifelse(counts>1 & true_count == 0, 1.2, count_percent))) %>% 
  filter(is_spike)

if (interactive()){
View(tmp2 %>% filter(is_spike) %>% filter(grepl("1_depth10000000_spike0.00000", sample), aligner == "bwa-mem", dbtype=="bins"))
View(tmp2 %>% filter(is_spike) %>% filter(grepl("1_depth10000000_spike0.10000", sample), aligner == "bwa-mem", dbtype=="bins"))
View(tmp2 %>% filter(is_spike) %>% filter(grepl("1_depth10000000_spike0.10000", sample), aligner == "bwa-mem", dbtype=="bins"))
}
(new <-ggplot(tmp2 %>% filter(is_spike) %>%
         group_by(target, aligner) %>%
         filter(!grepl("_masked", Genome)) %>% 
         mutate(xint = ifelse(row_number() == 1, true_count, NA)), aes(y=dbtype, x=counts, color=Genome)) +
  geom_vline(aes(xintercept = xint), color = "black", linetype = "dotted") + 
  geom_jitter(width=0) + 
  facet_grid( aligner ~  target, scales="free") + 
  theme(axis.text.x = element_text(angle=45, hjust=1),plot.caption = element_text(hjust = 0)
        ) + 
  scale_color_brewer(palette = "Paired")  +
  labs(caption="Dotted line indicate the targeted* of spike reads; x facet indicates the total spiked percentage of datase",
       subtitle="Version 2: more stringent alignment filtering")
)

if (FALSE) ggsave("old_v_new.pdf", width = 24, height=12, old+new+ plot_layout(ncol=2, guides = "collect"))
triple_pal <- c("#A6CEE3", "#1F78B4", "#47AFF4", "#B2DF8A", "#33A02C",  "#71C88A", "#FB9A99",  "#E31A1C","#FB7163")
ggplot(tmp2 %>% filter(is_spike) %>%
         group_by(target, aligner, genome_base) %>%
         mutate(xint = ifelse(row_number() == 1, true_count, NA)), aes(y=dbtype, x=counts, color=Genome)) +
  geom_vline(aes(xintercept = xint), color = "black", linetype = "dotted") + 
  geom_boxplot(outlier.colour = NA) + 
  geom_jitter(width=0, size =.1) + 
  facet_grid( aligner+genome_base ~  target, scales="free") + 
  theme(axis.text.x = element_text(angle=45, hjust=1),plot.caption = element_text(hjust = 0)
        ) + 
  scale_color_manual(values=triple_pal)  +
  labs(caption="Dotted line indicate the targeted* of spike reads; x facet indicates the total spiked percentage of datase",
       subtitle="Version 2: more stringent alignment filtering")



ggplot(tmp2 %>% filter(is_spike) %>%
         filter(target != 0) %>%  
         filter(Genome != "Trichoderma_reesei_primary") %>% 
         group_by(target, aligner, genome_base) %>%
         mutate(xint = ifelse(row_number() == 1, true_count, NA)), 
       aes(y=dbtype, x=counts/true_count * 100, color=Genome)) +
  geom_vline(xintercept = 100, color = "black", linetype = "dotted") + 
  geom_boxplot(outlier.colour = NA) + 
  geom_jitter(width=0, size =.1) + 
  facet_grid( aligner+genome_base ~  target, scales="free") + 
  theme(axis.text.x = element_text(angle=45, hjust=1),plot.caption = element_text(hjust = 0)
        ) + 
  scale_color_manual(values=triple_pal)  +
  labs(caption="Dotted line indicate the targeted* of spike reads; x facet indicates the total spiked percentage of datase",
       subtitle="Version 2: more stringent alignment filtering")



ggplot(tmp2 %>% filter(is_spike) %>% 
         filter(!is.infinite(count_percent)), aes(y=dbtype, x=count_percent, color=Genome)) +
  geom_jitter(width=0) + 
  facet_grid( aligner ~  target, scales="free") + 
  theme(axis.text.x = element_text(angle=45, hjust=1)) + scale_color_brewer(palette = "Paired")  

# if columns show the target and the organism
# scale across replicate, sligner quant ref, and dbtype
data_heat_summary <- tmp2 %>%
  filter(is_spike) %>% 
  group_by(sample, aligner,quantref, dbtype) %>%
  summarize(mx = max(count_diff), mn = min(count_diff),
            divby0 = (mx-mn) == 0) 

tmp2 %>% 
  filter(is_spike) %>% 
  inner_join(data_heat_summary) %>%
  mutate(p = (count_diff - mn ) / (mx - mn)) %>%
  ggplot(aes(x = genome_base, y = interaction(quantref, aligner, dbtype))) + 
  geom_tile(aes(fill = p), colour = "black") + 
  scale_fill_gradient2() + 
  facet_grid(aligner ~  target, space="free") +
  theme(axis.text.x = element_text(angle=45, hjust=1)) 
  theme(axis.text.x = element_text(angle = 30, hjust = 1)) +  geom_text(aes(label = count_diff)) 

  # here we only scale within a target type
data_heat_summary <- tmp2 %>%
  filter(is_spike) %>% 
  group_by(target, rep, genome_base) %>%
  summarize(mx = max(count_diff), mn = min(count_diff),
            divby0 = (mx-mn) == 0) 

tmp2 %>% 
  filter(rep==1) %>% 
  filter(is_spike) %>% 
  inner_join(data_heat_summary) %>%
  mutate(p = 2 * ((count_diff - mn ) / (mx - mn)) - 1) %>% # doesn't keep 0 in center
  mutate(p =  ifelse(count_diff < 0 , -count_diff[count_diff<0]/mn, count_diff[count_diff>=0]/mx  )) %>%
  ggplot(aes(x = interaction(rep , genome_base ), y = interaction(aligner, dbtype, quantref))) + 
  geom_tile(aes(fill = p), colour = "black") + 
  scale_fill_gradient2() + 
  facet_grid( ~  target, space="free") +
  theme(axis.text.x = element_text(angle=90, hjust=1))  +
  geom_text(aes(label = round(count_diff))) 

data_heat_summary <- tmp2 %>%
  filter(is_spike) %>% 
  group_by(target, genome_base, rep) %>%
  summarize(mx = max(count_diff), mn = min(count_diff)) 

tmp2 %>% 
  inner_join(data_heat_summary) %>%
  mutate(p = ((count_diff - mn ) / (mx - mn)) * sign(count_diff)) %>% 
  ggplot(aes(x = interaction(rep, aligner), y = interaction( dbtype, quantref))) + 
  geom_tile(aes(fill = p), colour = "black") + 
  scale_fill_gradient2() + 
  facet_grid(genome_base ~  target, space="free") +
  theme(axis.text.x = element_text(angle=90, hjust=1))  +
  geom_text(aes(label = round(count_diff))) + 
  coord_flip() + 
  labs(title="No major differences by aligner")


data_heat_summary <- tmp2 %>%
  filter(is_spike) %>% 
  filter(aligner == "bwa-mem") %>% 
  group_by(target, genome_base) %>%
  summarize(mx = max(count_diff), mn = min(count_diff)) 

tmp2 %>% 
  filter(aligner == "bwa-mem") %>% 
  inner_join(data_heat_summary) %>%
  mutate(p = ((count_diff - mn ) / (mx - mn)) * sign(count_diff)) %>% 
  ggplot(aes(y = interaction(rep, dbtype ), x = interaction(quantref))) + 
  geom_tile(aes(fill = p), colour = "black") + 
  scale_fill_gradient2(low = "#435F44", mid = "white", high = "firebrick") + 
  facet_grid(genome_base ~  target, space="free") +
  theme(axis.text.x = element_text(angle=90, hjust=1))  +
  geom_text(aes(label = round(count_diff))) +
  labs(x="Quantification Region", y = "Replicate and Alignment off-target database",
       title="Version2: filtered alignments")

```